Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida

### 2021-11-11 manu-draft.Rmd
Discussion tweaks
Mostly focused on changing context to 'release' impacts.

The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of the ride tide organism *K. brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.  The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  These conditions were widespread in lower and middle Tampa Bay.

In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters discharged from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).  From both a monitoring perspective and the nature of the discharge into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

Several of the results of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads. Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chl-a was first associated with a diatom bloom in April.  In June, a bloom of the red tide species *K. brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay through July. Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K. brevis* bloom (see supplement).  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread through a wide region of lower and middle Tampa Bay. Fisheries resources were impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].  Prior blooms were more localized than the 2021 bloom, potentially allowing fishes to migrate from harmful conditions.

For seagrasses, both major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

Macroalgae trends across the study period were much more dramatic than the minimal changes observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.  The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay. The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.  Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site. Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge. Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection. However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.

Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K. brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.  Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay. Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].  Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, also created conditions that were favorable for growth of *K. brevis*. However, the results suggest a likely scenario that residual nutrients from the Piney Point discharge, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K. brevis* to the concentrations observed in July [also see @Medina20].  Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the discharge was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production. Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the discharge.  Unlike Piney Point, inorganic nitrogen concentrations of the discharge were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the discharge stopped (see supplement).  The fate of the phosphorus in Tampa Bay is unknown.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the discharge, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.  Long-term fate of nutrients from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021. Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.  These efforts are critical to managing the health of coastal resources of the Gulf of Mexico given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.

>>>>

The observed conditions in Tampa Bay in 2021 following unanticipated releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of the ride tide organism *K. brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.  The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  These conditions were widespread in lower and middle Tampa Bay.

In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).  From both a monitoring perspective and the nature of the unanticipated release into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

Several of the results of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads. Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chl-a was first associated with a diatom bloom in April.  In April, the red tide species *K. brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay from June through July. Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K. brevis* bloom (see supplement).  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaked in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread throughout lower and middle Tampa Bay. Fisheries resources were impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall 2021.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].  Prior blooms were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions.

For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the initial study period. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

Macroalgae trends across the study period were much more dramatic than the minimal changes observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.  The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay. The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.  Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site. Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those in potential response to the Piney Point release. Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection. However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.

Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K. brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.  Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay. Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].  Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, also created conditions that were favorable for growth of *K. brevis*. However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K. brevis* to the concentrations observed in July [also see @Medina20].  Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the plume was widespread throughout the bay and persisted for many months after the release ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production. Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the water released.  Unlike Piney Point, inorganic nitrogen concentrations of the release were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).  The fate of the phosphorus in Tampa Bay is unknown.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the release, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.  Long-term fate of nutrients from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021. Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.  These efforts are critical to managing the health of coastal resources of the Gulf of Mexico given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
